---
title: "Predicting Weight-Lifting Exercise Quality from Accelerometer Data"
author: "Practical Machine Learning — Course Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4.2)
set.seed(1337)
```

# Overview

Wearable devices make it cheap to quantify *how much* activity a person does,
but rarely *how well* they do it. Using accelerometer, gyroscope and
magnetometer readings from sensors on the **belt, forearm, arm and dumbbell**
of six participants, this report builds a classifier that predicts the manner
in which a barbell lift was performed — the `classe` variable, with five
levels: exactly according to specification (**A**) and four common mistakes
(**B–E**).

The final model is a **random forest** trained on 52 sensor features. It
achieves an estimated **out-of-sample accuracy above 99%** (error < 1%),
validated both by 5-fold cross-validation and by a held-out validation set,
and it classifies all **20 quiz test cases**.

```{r libraries}
library(randomForest)
library(rpart)
library(rpart.plot)
```

# Loading the Data

The data originate from the Weight Lifting Exercise Dataset
(Velloso *et al.*, 2013). Strings `"NA"`, `""` and the Excel error
`"#DIV/0!"` are all read as missing values.

```{r load}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if (!file.exists("data/pml-training.csv")) {
  dir.create("data", showWarnings = FALSE)
  download.file(train_url, "data/pml-training.csv")
  download.file(test_url,  "data/pml-testing.csv")
}
na_strings <- c("NA", "", "#DIV/0!")
training <- read.csv("data/pml-training.csv", na.strings = na_strings)
testing  <- read.csv("data/pml-testing.csv",  na.strings = na_strings)
dim(training); dim(testing)
```

The raw data has **160 columns**. The target `classe` is reasonably balanced
across its five levels:

```{r classe-dist, fig.height=3.4}
training$classe <- factor(training$classe)
barplot(table(training$classe),
        col = "steelblue", border = NA,
        main = "Distribution of the outcome (classe)",
        xlab = "Exercise manner", ylab = "Count")
```

# Cleaning and Feature Selection

Two groups of columns carry no useful predictive signal and are removed:

1. **Summary/aggregate columns.** 100 columns (`kurtosis_*`, `skewness_*`,
   `max_*`, `min_*`, `avg_*`, `stddev_*`, `var_*`, `amplitude_*`) are populated
   only on the `new_window == "yes"` rows and are therefore *entirely missing*
   in the 20-case test set. A column that is all-`NA` in the test set cannot
   help predict it, so all near-empty columns are dropped.
2. **Identifier / bookkeeping columns.** The first seven columns
   (row index, `user_name`, three timestamps, `new_window`, `num_window`) are
   metadata. Using them would let the model "cheat" by memorising when each
   record was collected rather than learning the *physics* of the movement,
   which would not generalise to new subjects.

```{r clean}
# 1. drop columns that are mostly NA
na_frac  <- colMeans(is.na(training))
keep     <- na_frac < 0.90
training <- training[, keep]
testing  <- testing[,  keep]

# 2. drop the first 7 identifier / timestamp columns
training <- training[, -(1:7)]
testing  <- testing[,  -(1:7)]

dim(training)
```

This leaves **`r ncol(training) - 1` numeric sensor predictors** plus the
outcome — a complete-case matrix with no missing values remaining
(`sum(is.na) = `r sum(is.na(training))``).

# Data Splitting and Cross-Validation Strategy

To obtain an honest estimate of out-of-sample performance, the labelled data
is partitioned into a **70% training** set and a **30% validation** set. The
validation set is touched only once, at the very end, to estimate the
generalisation error.

```{r split}
n <- nrow(training)
idx      <- sample(seq_len(n), size = floor(0.70 * n))
train_df <- training[idx, ]
valid_df <- training[-idx, ]
c(train = nrow(train_df), validation = nrow(valid_df))
```

**Cross-validation** is used *inside* the training set to choose the model and
estimate its error before ever looking at the validation set. Below is an
explicit **5-fold cross-validation** of the random forest: the training data is
split into 5 folds; each fold is held out in turn while a forest is fit on the
other four, and accuracy is measured on the held-out fold.

```{r cv}
k      <- 5
folds  <- sample(rep(1:k, length.out = nrow(train_df)))
cv_acc <- numeric(k)
for (i in 1:k) {
  fit_i <- randomForest(classe ~ ., data = train_df[folds != i, ],
                        ntree = 150)
  pred_i <- predict(fit_i, train_df[folds == i, ])
  cv_acc[i] <- mean(pred_i == train_df$classe[folds == i])
}
round(cv_acc, 4)
cat(sprintf("Mean 5-fold CV accuracy: %.4f  (CV error estimate: %.4f)\n",
            mean(cv_acc), 1 - mean(cv_acc)))
```

The five folds agree closely (low variance), and the **mean CV error is well
under 1%**, so a random forest is clearly an appropriate choice.

# Model Building

## Why Random Forest

Random forests are a strong default for this problem: they capture the
non-linear interactions between joint angles and accelerations, are robust to
correlated predictors and different feature scales, need little tuning, and
provide a built-in **out-of-bag (OOB)** error estimate — itself a form of
cross-validation, since every tree is evaluated on the ~1/3 of samples it did
not see.

For context, a single decision tree (`rpart`) is fit first as an
interpretable baseline.

```{r tree, fig.height=4.6}
tree_fit <- rpart(classe ~ ., data = train_df, method = "class")
rpart.plot(tree_fit, type = 2, extra = 104, box.palette = "GnBu",
           tweak = 1.0, main = "Baseline: single decision tree")
tree_acc <- mean(predict(tree_fit, valid_df, type = "class") == valid_df$classe)
```

The single tree reaches only about **`r round(100*tree_acc)`% validation
accuracy** — decent, but far from enough. The random forest is fit next on the
full training set.

```{r rf}
rf_fit <- randomForest(classe ~ ., data = train_df, ntree = 250,
                       importance = TRUE)
rf_fit
```

The printed **OOB estimate of error rate** is under 1%, consistent with the
5-fold CV result above.

## Variable Importance

The most influential predictors are dominated by the belt sensor's roll/yaw
and the dumbbell/forearm readings — physically sensible, since these capture
the trunk and lifting-arm motion that distinguish correct from incorrect form.

```{r importance, fig.height=4.8}
varImpPlot(rf_fit, n.var = 15, type = 2,
           main = "Top 15 predictors (mean decrease in Gini)")
```

# Out-of-Sample Error Estimate

The model is now scored **once** on the held-out validation set — data it has
never seen — to estimate the true out-of-sample error.

```{r validate}
rf_pred <- predict(rf_fit, valid_df)
cm      <- table(Predicted = rf_pred, Actual = valid_df$classe)
cm
val_acc <- sum(diag(cm)) / sum(cm)
cat(sprintf("Validation accuracy: %.4f\n", val_acc))
cat(sprintf("Estimated out-of-sample error: %.4f (%.2f%%)\n",
            1 - val_acc, 100 * (1 - val_acc)))
```

**Expected out-of-sample error.** Three independent estimates agree:

| Method | Error estimate |
|---|---|
| 5-fold cross-validation | `r sprintf("%.2f%%", 100*(1-mean(cv_acc)))` |
| Random-forest OOB | `r sprintf("%.2f%%", 100*rf_fit$err.rate[nrow(rf_fit$err.rate),"OOB"])` |
| Held-out validation set | `r sprintf("%.2f%%", 100*(1-val_acc))` |

All three sit **below 1%**, so I expect the classifier to misclassify roughly
one lift in a hundred on new data from these participants. The agreement
between the OOB/CV estimates and the untouched validation set indicates the
model is **not overfit**.

# Predicting the 20 Test Cases

Finally, the fitted random forest is applied to the 20 quiz cases. To use all
available labelled data for the final predictions, the model is refit on the
full cleaned training set before predicting.

```{r predict-final}
final_rf <- randomForest(classe ~ ., data = training, ntree = 250)
final_predictions <- predict(final_rf, testing)
data.frame(problem_id = testing$problem_id, prediction = final_predictions)
```

Given the < 1% estimated per-observation error, the probability of getting all
20 predictions correct is roughly `0.99^20 ≈ 82%`, so a near-perfect quiz score
is expected.

# Conclusion

After removing metadata and the 100 all-missing summary columns, a random
forest on the 52 raw sensor features cleanly separates the five ways of
performing the lift. Cross-validation, the out-of-bag estimate, and an
independent validation set **all agree on an out-of-sample error below 1%**,
giving high confidence in the 20 test-case predictions above.

*Data source: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H.
"Qualitative Activity Recognition of Weight Lifting Exercises." Proceedings of
4th International Conference in Cooperation with SIGCHI (Augmented Human '13),
Stuttgart, Germany, 2013. Used with permission.*
