---
title: "mtcars MPG Predictor"
subtitle: "A Shiny app that estimates fuel economy"
author: "Developing Data Products -- Course Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## The problem

How many **miles per gallon** should you expect from a car, given a few basic
specs -- its weight, horsepower, cylinders and transmission?

- Buyers and students want a quick, interpretable estimate.
- The classic `mtcars` dataset (32 cars, 1974 *Motor Trend*) is perfect for a
  small, reproducible model.

The **mtcars MPG Predictor** turns that dataset into an interactive tool: move a
few sliders and get an instant, model-based MPG estimate.

**Live app:** `https://YOUR-USERNAME.shinyapps.io/mtcars-mpg-app/`
**Source:** `ui.R` and `server.R` on GitHub.

## The model (runs when slidified)

The app fits one ordinary least-squares regression on startup. Here it is,
evaluated live in this slide deck:

```{r model}
fit <- lm(mpg ~ wt + hp + factor(cyl) + factor(am), data = mtcars)
round(coef(fit), 2)
```

Interpretation: heavier, higher-horsepower cars lose MPG; a manual transmission
(`factor(am)1`) and fewer cylinders tend to add MPG. The model explains a large
share of the variation:

```{r rsq}
paste0("Adjusted R-squared: ", round(summary(fit)$adj.r.squared, 3))
```

## A worked prediction

A user who dials in a 3,200-lb, 150-hp, 6-cylinder car with a manual gearbox
gets this estimate -- computed exactly as the Shiny server does:

```{r predict}
newcar <- data.frame(wt = 3.2, hp = 150, cyl = 6, am = 1)
round(predict(fit, newcar), 1)
```

```{r plot, echo=FALSE, fig.height=3.2, fig.width=9}
library(ggplot2)
newcar$mpg <- predict(fit, newcar)
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(color = "steelblue", size = 2, alpha = 0.7) +
  geom_point(data = newcar, aes(wt, mpg), color = "red", size = 5) +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon",
       title = "Your car (red) vs. the 32 mtcars cars") +
  theme_minimal(base_size = 13)
```

## Using the app

The interface is deliberately simple -- no R knowledge required:

1. **Sliders** set weight and horsepower.
2. **Radio buttons** choose cylinders (4 / 6 / 8) and transmission.
3. The predicted **MPG** updates instantly at the top of the page.
4. A **Plotly** chart marks your car in red against the real data.

Every screen has built-in help, and a **Documentation** tab explains the model
in plain language. Inputs flow through a `reactive()` expression in `server.R`,
so output refreshes with no button press.

## Why it's useful

- **Reproducible:** one dataset, one transparent linear model, all in ~90 lines
  of `ui.R` + `server.R`.
- **Interactive & instant:** reactive prediction and an interactive Plotly plot.
- **Self-documenting:** novice users get started with no prerequisites.
- **Extensible:** swap in another dataset or add predictors in a few lines.

> Try it, move the sliders, and see how weight and horsepower trade off against
> fuel economy.

Thank you!
